\doctitle

### 2023-06-08 pinode.tex
[FINAL SUBMITTED] The very final version of my thesis. Compiles clean.

\section{Introduction}
\subsection{Related Work}
        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \cite{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \cite{farahmand2016learning,nabi2022robust}. At the time of writing this document, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \cite{otterness2017evaluation}. Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \cite{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

        In order to enable control of high dimensional dynamical systems, a ROM training method needs to  identify a low-dimensional manifold along with dynamics on the manifold that together yield high-accuracy predictions and long-term stability \cite{ahmed2021closures,noack2011reduced}. Most traditional ROMs are projection-based, e.g. dynamic mode decomposition (DMD) \cite{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \cite{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace. This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of the basis functions (spatial modes) \cite{benner2015survey}. One challenge for POD methods is their intrusive nature, i.e. requiring access to the solver codes. To overcome this, operator inference approaches \cite{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and  exploit lifting to fit the latent space dynamics data into polynomial, typically quadratic, models. These models, however, are (i) limited in representation power (up to quadratic, e.g. for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.

        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \cite{lee2020model,gin2021deep,champion2019data,kim2019deep}. In practice, however, autoencoder-based ROMs require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system. Moreover, the large demand for training data significantly limits the use of such models in physics applications where the data can be expensive to obtain.

        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \cite{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs. For example, in HVAC applications, one may collect data from a room with two windows but not from a room for every possible number of windows. In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \cite{nabi2020improving}. In such situations embedding the knowledge of physics into a model becomes necessary to improve the extrapolation performance, and for which several approaches have recently been proposed.
        For instance, the seminal works \cite{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression. Recently, Cranmer et al. \cite{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations. They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution data better than the GNN itself. However, symbolic regression also suffers from excessive computational costs,  and may be prone to overfitting.

        Another example of incorporating physics in ROMs is the use of parametric models at the latent space, e.g. by using the sparse identification of nonlinear dynamics (SINDy) \cite{brunton2016discovering,champion2019data}. For instance, \cite{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics. However, such loss is highly sensitive to noise in the data, especially when evaluating time-derivatives with finite differences is required\cite{delahunt2022sindynoise}. 
        Collocation-based enforcement of the physics, i.e. projection of the candidate functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties. Recently, Liu et al. \cite{liu2022physics} used an auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM. However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \cite{page2019koopman}.
        Finally, recent works on NeuralODE (NODE) \cite{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g. a network) as a latent space dynamics model, significantly extending the set of models for the latent dynamics that one can train efficiently.

@@

In real-world applications, it is often expensive to solve equation~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional. However, a variety of works provided both theoretical \cite{holmes2012turbulence} and empirical \cite{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$. In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:

@@

Multiple real-world applications necessitate using ROMs instead of integrating the relation~(\ref{eq:generic_stationary_ode}) directly. For example, integrating~(\ref{eq:generic_stationary_ode}) may be computationally intractable especially on platforms with limited computing capability such as embedded and autonomous devices. For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stokes equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances. On the other hand, integrating~(\ref{eq:integration_in_latent_space}) may be cheap when $m << n$. Finally, even when solving~(\ref{eq:generic_stationary_ode}) is possible in real time (e.g. by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable. Indeed, executing control requires \textit{multiple} evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\cite{duriez2017machine}.

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/abstract_data_driven.png}
    \caption{Illustration of the autoencoer structure with neural ODE in the latent space. The data-driven part of the loss function aims to minimize a sum of two objectives: the prediction loss and the reconstruction loss. The prediction loss minimizes the difference between the data trajectories and their model predictions to ensure temporal consistency of the latent space dynamics. The reconstruction loss ensures accurate reconstruction of individual snapshots, ensuring that the autoencoder behaves as an invertible mapping on all snapshots.}
    \label{fig:data_driven_loss}
\end{figure}

@@

\paragraph{Data-Driven Loss} Similar to prior works \cite{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses. The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data, as illustrated on Figure~\ref{fig:data_driven_loss}.

@@

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\cite{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation, without loss of generality, in the rest of the section we assume that all trajectories are recorded over the same time-frame with the same uniform step-size. To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs)\cite{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}). Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited. This is especially useful in applications where the size of the latent space dimension needs to be small~\cite{lee2020model,gin2021deep,champion2019data,kim2019deep}.

\paragraph{Physics-Informed Loss} In their recent work, Liu et al.~\cite{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics. To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
\begin{equation}
    \label{eq:chain_rule_1}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bd{f}(\bd{x(t)})
\end{equation}
On the other hand, by the definition of $\psi$ and $\bd{h}$ we have that
\begin{equation}
    \label{eq:chain_rule_2}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \bd{h}(\phi(\bd{x}(t))
\end{equation}
Combining Equations~(\ref{eq:chain_rule_1}) and~(\ref{eq:chain_rule_2}) we get that
\begin{equation}
    \label{eq:physsics_informed_equation}
    \bd{h}(\phi(\bd{x}(t)) = \nabla\phi(\bd{x})^T\bd{f}(\bd{x})
\end{equation}

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/abstract_physics_informed.png}
    \caption{The physics-informed loss function compares gradient fields in the current latent space with what a correctly-learned field should be in this latent space on set of collocation points.}
    \label{fig:physics_informed_loss}
\end{figure}

@@

The above definition of collocation points is not to be confused with a classic notion of collocation points for finding numerical solutions for differential equations~\cite{fornberg1998practical, trefethen2022numerical}. The classic notion refers to a set of points in time $[t_0, t_0 + c_1h, t_0 + c_2h, \dots, t_0 + h]$, $0 < c1 < c2 < \dots < 1$ which are chosen to obtain an optimal local interpolant of a solution of a differential equation for a time-period between $t_0$ and $t_0 + h$. For example, $s$ collocation points for Runge-Kutta methods are defined to provide an optimal Gauss-Legendre interpolant of order $s$; the coefficients $c_1, \dots, c_s$ come from a respective Butcher table. In contrast, we define collocation points as pairs $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$ which are  examples of mapping $x \to f(x)$. Our definition is built around solving an \textit{inverse} problem of approximating $\dot{x} = f(x)$ with $f_\theta(x)$ and follows a recent work~\cite{liu2022physics} which develops upon a definition from \cite{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.

@@

We use a \texttt{pytorch}~\cite{NEURIPS2019_9015} implementation of the Adam algorithm~\cite{kingma2014adam} for optimization. To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\cite{chen2018neural} -- a \texttt{pytorch}-compatible implementation of the Neural ODE framework.

To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), autoencoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}). Thus, we call our framework \textit{Physics-Informed Neural ODE}, or PINODE. 
\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/duff_first_exp_abstract.png}
    \caption{We use a toy example -- a Lifted Duffing Oscillator -- to show that it is possible to ``fill the gaps'' in data with collocation points. Specifically, the Hybrid model is able to learn the dynamics of two additional basins of attraction that were not represented in the dataset. As shown in the top-rightmost frame, without the collocation points the model does not infer the dynamics in the unseen regions correctly.}
    \label{fig:duffing_pinode}
\end{figure}

\section{Experiments}
\label{sec:exp}
The experiments section is organized as follows. First, to illustrate the ideas behind the framework we study its performance on a high-dimensional ODE -- a lifted Duffing oscillator. We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \cite{liu2022physics} by handling multiple basins of attraction within one model. We also show that using physics-informed loss is sufficient for reconstructing the behaviour for basins of attraction that are not represented by the data. Finally, we demonstrate that a purely data-driven model may be highly-accurate in the short-term and highly unstable in the long-term, even when the data is abundant, and show that the physics-informed approach improves long-term stability of such models by multiple orders of magnitude.

@@

\begin{wrapfigure}{r}{0.25\textwidth}
  \begin{center}
    \includegraphics[width=0.23\textwidth]{figures/duffing_comparison.pdf}
  \end{center}
  \caption{Non-linearity in the latent dynamics and the autoencoder employed in teh PINODE Hybrid model are important for accurate long-term extrapolation. The DMD model and PIKN Hybrid model were unable to extrapolate the dynamics from collocation points.}%\vspace{-0.5in}
  \label{fig:duffing_comparison}
\end{wrapfigure}
A phase portrait for 300 randomly sampled trajectories from this system is visualized on Figure~\ref{fig:duffing_pinode}, left frame. Depending on the total energy, each trajectory always stays in one of three regions: the left lobe, the right lobe, or the outer area, visualized in red, green, and blue, respectively. To create a synthetic high-dimensional system that retains this property, we lift the Duffing trajectories into a higher-dimensional space by applying an invertible transformation $\mathcal{A}(\bd{z})$:

@@

Hence, for this system $z \in \ZZ = \mathbb{R}^2$ and $\bd{x} \in \XX = \text{span}\{A_{:,1}, A_{:,2}\} \subseteq \mathbb{R}^{128}$. We treat $\XX$ as an observable space, in which the dynamical system~(\ref{eq:duffing_definition}) obeys the following:
\begin{equation}
    \frac{d\bd{x}}{dt} = \bd{f}(\bd{x}) = \nabla((A^TA)^{-1}A^T\bd{x}^{1/3})^T\bd{h}((A^TA)^{-1}A^T\bd{x}^{1/3})
\end{equation}
Thus, we created a high-dimensional dynamical system with multiple basins of attraction for which the dynamics $\bd{f}$ are known.

@@

We make two observations from the results displayed in Figure~\ref{fig:duffing_pinode}. First, a purely data-driven model is unable to extrapolate outside its training region using only the data from that region. This observation is consistent with the conclusions from related works~\cite{gin2021deep} that neural networks interpolate well but struggle with extrapolation tasks. Second, we see that collocation points provided enough extra information for the model to predict nearly perfectly in regions from which no trajectories were provided. This observation suggests that one can use collocation points to ``cover the gaps'' in data and improve the extrapolation accuracy of the model. 
\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/duffing_periods.pdf}
    \caption{ Box plots of the prediction error for three PINODE models: Data-Driven, Physics-Informed, and Hybrid. The time is measured in multiples of the training time period, i.e. $x=3T$ refers the time-range between two and three training time-periods away. }%\vspace{-0.4in}
    \label{fig:duffing_periods}
\end{figure}

The ability of Neural ODE to model nonlinear dynamics in the latent space is demonstrated in Figure~\ref{fig:duffing_comparison}. The figure shows a comparison between the Hybrid PINODE model, the Hybrid PIKN model~\cite{liu2022physics}, and DMD, all of which have been trained using the same dataset. PIKN differs from PINODE in that it uses linear latent dynamics $\frac{dz}{dt} = Lz$, where $L$ is a finite-dimensional approximation of the Koopman operator, instead of a general non-linear dynamics operator $\frac{dz}{dt} = h_\theta(z)$. For PIKN, we set $z \in \mathbb{R}^{16}$, an 8 times expansion of the dimension of the true manifold. We observe in Figure~\ref{fig:duffing_comparison} that PIKN is unable to extrapolate the dynamics to unseen areas correctly using the collocation points: eventually, all trajectories "collapse" onto the same attractor. It can also be seen that DMD shows even worse performance which could be attributed to its linear model reduction.

@@

\label{sec:compressibility}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/compressibility.pdf}
    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently in terms of MSE than PIKN Hybrid model when modelling low-viscosity (highly-nonlinear) Burgers' equation (left frame). The difference in performance grows to x100, when forecasting two times farther than the training period (central frame). PIKN suffers from long-term instability due to the presence of eigenvalues with positive real part in the latent dynamics matrix (right frame). In this frame we plot all the eigenvalues of the latent-space matrix for each PIKN model from frames 1-2. The legend in the right frame refers to the dimension of the latent space used by the corresponding PIKN model.}
    \label{fig:burgers_compressibility}
\end{figure}

>>>>

\section{Introduction}
\subsection{Related Work}
        Forecasting the behavior of a large-scale real-world system directly from first principles often requires solving highly-nonlinear governing equations such as high-dimensional ordinary differential equations (ODEs) or partial differential equations (PDEs). 
        High-fidelity simulations of such dynamical systems can become intractable, especially if an online control algorithm requires multiple forecasts per second using a low-powered embedded device \citep{rowley2017model,lucia2004reduced,benner2015survey}. A situation like this arises, for example, when a smart heating, ventilation, and air conditioning (HVAC) system attempts to optimize the temperature distribution of the air in a room using only partial measurements \citep{farahmand2016learning,nabi2022robust}. At the time of writing this document, such systems are incapable of real-time complex simulations, but they can already run low-dimensional pre-trained models, which invites the development of high-quality reduced order models (ROMs) \citep{otterness2017evaluation}. Therefore, ROMs are essential for enabling the design optimization, uncertainty propagation, predictive modeling, and control for such dynamical systems \citep{brunton2022data,kutz2016dynamic,rowley2017model,jones2020characterising}

        In order to enable control of high dimensional dynamical systems, a ROM training method needs to  identify a low-dimensional manifold along with dynamics on the manifold that together yield high-accuracy predictions and long-term stability \citep{ahmed2021closures,noack2011reduced}. Most traditional ROMs are projection-based, e.g. dynamic mode decomposition (DMD) \citep{kutz2016dynamic,tu2013dynamic} and proper orthogonal decomposition (POD) \citep{holmes2012turbulence}, which transform the trajectories of a high-dimensional dynamical system into a suitable, and in some sense optimal, low-dimensional subspace. This projection leads to truncation of higher order modes and parametric uncertainties, which result in large prediction errors over time due to the deterioration of the basis functions (spatial modes) \citep{benner2015survey}. One challenge for POD methods is their intrusive nature, i.e. requiring access to the solver codes. To overcome this, operator inference approaches \citep{qian2020lift,peherstorfer2016data} utilize SVD-based model reduction and  exploit lifting to fit the latent space dynamics data into polynomial, typically quadratic, models. These models, however, are (i) limited in representation power (up to quadratic, e.g. for lift and learn approach) and (ii) require a custom-tailored SVD-based optimization technique.

        In a thrust to overcome these challenges, significant effort has been invested into developing autoencoder-based reduced-order models, as a popular nonlinear ROM technique, which can yield both accurate and stable ROMs \citep{lee2020model,gin2021deep,champion2019data,kim2019deep}. In practice, however, autoencoder-based ROMs require datasets that densely cover a hypothetical infinite dimensional phase portrait of the dynamical system. Moreover, the large demand for training data significantly limits the use of such models in physics applications where the data can be expensive to obtain.

        Another severe challenge of utilizing ROMs comes from their poor out-of-distribution performance \citep{fries2022lasdi,cranmer2020discovering,gin2021deep}, especially when it is fundamentally impossible for a practitioner to obtain data that covers the entire distribution of possible data inputs. For example, in HVAC applications, one may collect data from a room with two windows but not from a room for every possible number of windows. In atmospheric LiDAR applications, we may conduct experiments on a certain terrain but we can never conduct experiments on all sorts of terrains \citep{nabi2020improving}. In such situations embedding the knowledge of physics into a model becomes necessary to improve the extrapolation performance, and for which several approaches have recently been proposed.
        For instance, the seminal works \citep{bongard2007automated,schmidt2009distilling} have tried to determine the underlying structure of a nonlinear dynamical system from data using symbolic regression. Recently \citep{cranmer2020discovering} employed symbolic regression in conjunction with graph neural network (GNN), while encouraging sparse latent representation, to extract explicit physical relations. They showed that the symbolic expressions extracted from the GNN generalized to out-of-distribution data better than the GNN itself. However, symbolic regression also suffers from excessive computational costs,  and may be prone to overfitting.

        Another example of incorporating physics in ROMs is the use of parametric models at the latent space, e.g. by using the sparse identification of nonlinear dynamics (SINDy) \citep{brunton2016discovering,champion2019data}. For instance, \citep{fries2022lasdi,he2022glasdi} used a chain-rule based loss that ties latent-space derivatives to the observable-space derivatives for simultaneous training of the auto-encoder and the latent dynamics. However, such loss is highly sensitive to noise in the data, especially when evaluating time-derivatives with finite differences is required\citep{delahunt2022sindynoise}. 
        Collocation-based enforcement of the physics, i.e. projection of the candidate functions in the governing equations to enforce the chain rule instead of finite difference, could address such numerical difficulties. Recently, Liu et al. \citep{liu2022physics} used an auto-encoder architecture and Koopman theory to demonstrate that combining autoencoders with enforcing linear dynamics in the latent space may result in an interpretable ROM. However, linearity may not be expressive enough for complex dynamics with multiple basins of attraction \citep{page2019koopman}.
        Finally, recent works on NeuralODE (NODE) \citep{chen2018neuralode,rackauckas2020udes} show a way to fit an arbitrary non-linear model (e.g. a network) as a latent space dynamics model, significantly extending the set of models for the latent dynamics that one can train efficiently.

@@

In real-world applications, it is often expensive to solve equation~(\ref{eq:generic_stationary_ode}) directly because $x(t)$ can be very high-dimensional. However, a variety of works provided both theoretical \citep{holmes2012turbulence} and empirical \citep{noack2011reduced,chen2021discovering} evidence that many physical systems evolve on a manifold $\ZZ \subseteq \mathbb{R}^m$ of a lower dimension $m << n$. In that space, the dynamics evolve according to a (generally unknown) function  $\bd{h}(\bd{z})$:

@@

Multiple real-world applications necessitate using ROMs instead of integrating the relation~(\ref{eq:generic_stationary_ode}) directly. For example, integrating~(\ref{eq:generic_stationary_ode}) may be computationally intractable especially on platforms with limited computing capability such as embedded and autonomous devices. For instance, in an HVAC system, solving~(\ref{eq:generic_stationary_ode}) means solving a Navier-Stokes equation on a fine grid in real time, which exceeds the computing capabilities of current-generation appliances. On the other hand, integrating~(\ref{eq:integration_in_latent_space}) may be cheap when $m << n$. Finally, even when solving~(\ref{eq:generic_stationary_ode}) is possible in real time (e.g. by utilizing a remote cluster), executing control over the resulting model, which is an end-goal for an HVAC system, may still be intractable. Indeed, executing control requires \textit{multiple} evaluations of~(\ref{eq:generic_stationary_ode}) for \textit{each} iteration of control even for the most efficient algorithms known to date~\citep{duriez2017machine}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/abstract_data_driven.png}
    \caption{Illustration of the autoencoer structure with neural ODE in the latent space. The data-driven part of the loss function aims to minimize a sum of two objectives: the prediction loss and the reconstruction loss. The prediction loss minimizes the difference between the data trajectories and their model predictions to ensure temporal consistency of the latent space dynamics. The reconstruction loss ensures accurate reconstruction of individual snapshots, ensuring that the autoencoder behaves as an invertible mapping on all snapshots.}
    \label{fig:data_driven_loss}
\end{figure}

@@

\paragraph{Data-Driven Loss} Similar to prior works \citep{takeishi2017learning,morton2019deep,gin2021deep}, we define a \textit{data-driven loss} $\LL_{data}$ as a sum of reconstruction and prediction losses. The former ensures that $\phi_\theta$ and $\psi_\theta$ are inverse mappings of each other, whereas the latter matches the model's predictions to the available data, as illustrated on Figure~\ref{fig:data_driven_loss}.

@@

\begin{align}
    \label{eq:loss_data_driven}
    \LL^{data}_\theta & = \frac{1}{2\sigma^2}\sum_{i = 1}^k \left[\frac{\omega_1}{p}\sum_{j=1}^p\left\|\bd{x_i}(t_j) - \psi_\theta(\phi_\theta(\bd{x_i}(t_j)))\right\|^2\right. + \\
     & + \left.\frac{\omega_2}{p}\sum_{j=1}^p \left\|\psi_\theta\left(\phi_\theta(\bd{x_i}(t_1)) + \int_{t_1}^{t_j}h(z(t))dt\right) - \bd{x_i}(t_j)\right\|^2 \right]
\end{align}
where $\sigma$ is the standard deviation of the observation noise. We note that each trajectory $\bd{x}_i$ may be captured over its own time-frame and may use a distinct, possibly non-uniform, step-size, in which case the loss function should be modified accordingly\footnote{The implementation is affected only in evaluating the integral in~(\ref{eq:loss_data_driven}). This part is handled by \texttt{torchdiffeq}~\citep{chen2018neural} library, which supports non-uniform time-frames within a batch}. To simplify the notation, without loss of generality, in the rest of the section we assume that all trajectories are recorded over the same time-frame with the same uniform step-size. To forecast the behavior of the system in the latent space, we apply the technique of Neural Ordinary Differential Equations (Neural ODEs or NODEs by~\citep{chen2018neuralode}, which utilizes the adjoint sensitivity method to back-propagate the gradients through the integral in~(\ref{eq:loss_data_driven}). Neural ODEs have demonstrated a better ability to model highly non-linear dynamics compared to linear models when the dimensionality of the dynamics variable is limited. This is especially useful in applications where the size of the latent space dimension needs to be small~\citep{lee2020model,gin2021deep,champion2019data,kim2019deep}.

\paragraph{Physics-Informed Loss} In their recent work, Liu et al.~\citep{liu2022physics} proposed a method for utilizing knowledge of the governing equations $d\bd{x}/dt = \bd{f(x)}$ as a finite-dimensional approximation of Koopman eigenfunctions for linear latent dynamics. To extend this approach to the non-linear regime, we note that for a true mapping $\phi$ the following holds:
\begin{equation}
    \label{eq:chain_rule_1}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \frac{d\bd{z}}{d\bd{x}}\frac{d\bd{x}}{dt} = \nabla\phi(\bd{x}(t))^T\bd{f}(\bd{x(t)})
\end{equation}
On the other hand, by the definition of $\psi$ and $\bd{h}$ we have that
\begin{equation}
    \label{eq:chain_rule_2}
    \frac{d\bd{z}(\bd{x}(t))}{dt} = \bd{h}(\phi(\bd{x}(t))
\end{equation}
Combining Equations~(\ref{eq:chain_rule_1}) and~(\ref{eq:chain_rule_2}) we get that
\begin{equation}
    \label{eq:physsics_informed_equation}
    \bd{h}(\phi(\bd{x}(t)) = \nabla\phi(\bd{x})^T\bd{f}(\bd{x})
\end{equation}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/abstract_physics_informed.png}
    \caption{The physics-informed loss function compares gradient fields in the current latent space with what a correctly-learned field should be in this latent space on set of collocation points.}
    \label{fig:physics_informed_loss}
\end{figure}

@@

The above definition of collocation points is not to be confused with a classic notion of collocation points for finding numerical solutions for differential equations~\citep{fornberg1998practical, trefethen2022numerical}. The classic notion refers to a set of points in time $[t_0, t_0 + c_1h, t_0 + c_2h, \dots, t_0 + h]$, $0 < c1 < c2 < \dots < 1$ which are chosen to obtain an optimal local interpolant of a solution of a differential equation for a time-period between $t_0$ and $t_0 + h$. For example, $s$ collocation points for Runge-Kutta methods are defined to provide an optimal Gauss-Legendre interpolant of order $s$; the coefficients $c_1, \dots, c_s$ come from a respective Butcher table. In contrast, we define collocation points as pairs $(\bd{\bar{x}},\, \bd{f}(\bd{\bar{x}}))$ which are  examples of mapping $x \to f(x)$. Our definition is built around solving an \textit{inverse} problem of approximating $\dot{x} = f(x)$ with $f_\theta(x)$ and follows a recent work~\citep{liu2022physics} which develops upon a definition from \citep{raissi2018hidden} with the difference being the sample space: instead of sampling from the spatiotemporal domain we sample them from an appropriate function space.

@@

We use a \texttt{pytorch}~\citep{NEURIPS2019_9015} implementation of the Adam algorithm~\citep{kingma2014adam} for optimization. To evaluate $\nabla_\theta \LL^{physics}_\theta$ and $\nabla_\theta \LL^{data}_\theta$ we use \texttt{torchdiffeq}~\citep{chen2018neural} -- a \texttt{pytorch}-compatible implementation of the Neural ODE framework.

To the best of our knowledge, this is the first framework that combines non-linear latent-dynamics (Neural ODE), autoencoders, and a physics-informed loss term~(\ref{eq:loss_physics_informed}). Thus, we call our framework \textit{Physics-Informed Neural ODE}, or PINODE. 
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/duff_first_exp_abstract.png}
    \caption{We use a toy example -- a Lifted Duffing Oscillator -- to show that it is possible to ``fill the gaps'' in data with collocation points. Specifically, the Hybrid model is able to learn the dynamics of two additional basins of attraction that were not represented in the dataset. As shown in the top-rightmost frame, without the collocation points the model does not infer the dynamics in the unseen regions correctly.}
    \label{fig:duffing_pinode}
\end{figure}

\section{Experiments}
\label{sec:exp}
The experiments section is organized as follows. First, to illustrate the ideas behind the framework we study its performance on a high-dimensional ODE -- a lifted Duffing oscillator. We show how a non-linear latent dynamics $\bd{h}(\bd{z})$ overcomes the limitations of DMD and Koopman networks from \citep{liu2022physics} by handling multiple basins of attraction within one model. We also show that using physics-informed loss is sufficient for reconstructing the behaviour for basins of attraction that are not represented by the data. Finally, we demonstrate that a purely data-driven model may be highly-accurate in the short-term and highly unstable in the long-term, even when the data is abundant, and show that the physics-informed approach improves long-term stability of such models by multiple orders of magnitude.

@@

A phase portrait for 300 randomly sampled trajectories from this system is visualized on Figure~\ref{fig:duffing_pinode}, left frame. Depending on the total energy, each trajectory always stays in one of three regions: the left lobe, the right lobe, or the outer area, visualized in red, green, and blue, respectively. To create a synthetic high-dimensional system that retains this property, we lift the Duffing trajectories into a higher-dimensional space by applying an invertible transformation~$\mathcal{A}(\bd{z})$:

\begin{figure}
  \includegraphics[width=\textwidth]{figures/duffing_comparison.pdf}
  \caption{\label{fig:duffing_comparison}Non-linearity in the latent dynamics and the autoencoder employed in teh PINODE Hybrid model are important for accurate long-term extrapolation. The DMD model and PIKN Hybrid model were unable to extrapolate the dynamics from collocation points.}%\vspace{-0.5in}
\end{figure}

@@

Hence, for this system $z \in \ZZ = \mathbb{R}^2$ and $\bd{x} \in \XX = \text{span}\{A_{:,1}, A_{:,2}\} \subseteq \mathbb{R}^{128}$. We treat $\XX$ as an observable space, in which the dynamical system~(\ref{eq:duffing_definition}) obeys the following:
\begin{equation}
    \frac{d\bd{x}}{dt} = \bd{f}(\bd{x}) = \nabla((A^TA)^{-1}A^T\bd{x}^{1/3})^T\bd{h}((A^TA)^{-1}A^T\bd{x}^{1/3})
\end{equation}

Thus, we created a high-dimensional dynamical system with multiple basins of attraction for which the dynamics $\bd{f}$ are known.

@@

We make two observations from the results displayed in Figure~\ref{fig:duffing_pinode}. First, a purely data-driven model is unable to extrapolate outside its training region using only the data from that region. This observation is consistent with the conclusions from related works~\citep{gin2021deep} that neural networks interpolate well but struggle with extrapolation tasks. Second, we see that collocation points provided enough extra information for the model to predict nearly perfectly in regions from which no trajectories were provided. This observation suggests that one can use collocation points to ``cover the gaps'' in data and improve the extrapolation accuracy of the model. 
\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/duffing_periods.pdf}
    \caption{ Box plots of the prediction error for three PINODE models: Data-Driven, Physics-Informed, and Hybrid. The time is measured in multiples of the training time period, i.e. $x=3T$ refers the time-range between two and three training time-periods away. }%\vspace{-0.4in}
    \label{fig:duffing_periods}
\end{figure}

The ability of Neural ODE to model nonlinear dynamics in the latent space is demonstrated in Figure~\ref{fig:duffing_comparison}. The figure shows a comparison between the Hybrid PINODE model, the Hybrid PIKN model~\citep{liu2022physics}, and DMD, all of which have been trained using the same dataset. PIKN differs from PINODE in that it uses linear latent dynamics $\frac{dz}{dt} = Lz$, where $L$ is a finite-dimensional approximation of the Koopman operator, instead of a general non-linear dynamics operator $\frac{dz}{dt} = h_\theta(z)$. For PIKN, we set $z \in \mathbb{R}^{16}$, an 8 times expansion of the dimension of the true manifold. We observe in Figure~\ref{fig:duffing_comparison} that PIKN is unable to extrapolate the dynamics to unseen areas correctly using the collocation points: eventually, all trajectories "collapse" onto the same attractor. It can also be seen that DMD shows even worse performance which could be attributed to its linear model reduction.

@@

\label{sec:compressibility}
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/compressibility.pdf}
    \caption{PINODE Hybrid model utilized the latent space dimension 5 times more efficiently in terms of MSE than PIKN Hybrid model when modelling low-viscosity (highly-nonlinear) Burgers' equation (left frame). The difference in performance grows to x100, when forecasting two times farther than the training period (central frame). PIKN suffers from long-term instability due to the presence of eigenvalues with positive real part in the latent dynamics matrix (right frame). In this frame we plot all the eigenvalues of the latent-space matrix for each PIKN model from frames 1-2. The legend in the right frame refers to the dimension of the latent space used by the corresponding PIKN model.}
    \label{fig:burgers_compressibility}
\end{figure}
